Cybersecurity impacts at the beginning\\ of Covid-19 pandemic in Italy

### 2021-03-08 article.tex
Minor fix
Minor changes

In the pandemic period between 2019 and 1st quarter of 2020, the attacks on IAA that aimed at access control systems are most widespread with 34\% in 2019 and 28.2\% (Fig.~\ref{fig:5}).
As diffusion among the respondents, the other 14 types of attacks follow, decreasing by a few percentage points among them, whose characteristics and impacts are detailed in the related 
paragraphs of the OAD 2020 report. Regarding attack vectors and techniques, all the seven attacks families considered in the 2020 questionnaire are widely used in the various 
types of attacks, some times even simultaneously, as shown in Fig.~\ref{fig:6}. The trend of Fig.~\ref{fig:6} also mirrors for the most part the european threads recorded by European Network and Information Security Agency (ENISA) 
in the same period~\cite{enisa20}. The most widespread one, as an average on the various attacks detected by the respondents’ pool, is the use of toolkits (rootkits, meta-exploits, etc.) for the identification and exploitation of vulnerabilities
on the target system of the attack, with a 38.8\%, followed by the well-known malicious and unauthorized collection of information (social engineering, phishing, etc.) with 34.6\% 
and the use of malicious code and scripts with 34.3\%. The other considered technologies decrease with a few percentage points of difference between them. The impacts of the most 
critical attacks are analyzed for each attack type, as well as their possible motivations and the recovery times required by the most critical ones.
In the 2020 OAD survey, all these attacks’ characteristics vary for each type of attack, and the emerged results are described in the specific paragraphs dedicated to each attack 
type. At overall it emerges that:
\begin{itemize}
\item the impacts declared by the respondents are balanced between those irrelevant and/or easily resolvable with recovery in short time and at limited costs, and those very 
critical, which require expensive and long recovery actions, and that, in some cases, can cause business and customer loses; these two very different impact cases depend mainly 
on the security measures in place;

@@

In particular, the European Union Agency For Cybersecurity(ENISA)

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig5.png}
		\caption{Diffusion of digital attacks on respondents' computer systems based on the type of attack between 2019 - 1st quoter 2020 (multiple answers) \cite{oad20}}
		\label{fig:5}
\end{figure}
\begin{figure}
	\centering
		\includegraphics[width=0.90\textwidth]{pictures/fig6.png}
		\caption{Average diffusion (percentage) of the attack techniques detected in the most serious attacks between 2019 - 1st quoter 2020 (multiple answers) \cite{oad20}}
		\label{fig:6}
\end{figure}

>>>>

In the pandemic period between 2019 and 1st quarter of 2020,  the nature of attacks and their vectors were quite diversified.
The attacks on IAA that aimed at access control systems are most widespread with 34\% in 2019 and 28.2\% (Fig.~\ref{fig:5}).

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig5.png}
		\caption{Diffusion of digital attacks on respondents' computer systems based on the type of attack between 2019 - 1st quoter 2020 (multiple answers) \cite{oad20}}
		\label{fig:5}
\end{figure}

As diffusion among the respondents, the other 14 types of attacks follow, decreasing by a few percentage points among them, whose characteristics and impacts are detailed in the related 
paragraphs of the OAD 2020 report. Regarding attack vectors and techniques, all the seven attacks families considered in the 2020 questionnaire are widely used in the various 
types of attacks, some times even simultaneously, as shown in Fig.~\ref{fig:6}. The trend of Fig.~\ref{fig:6} is also in line with for the most part the european threads recorded in ~\cite{enisa20} by European Network and Information Security Agency (ENISA). 
The most widespread one, as an average on the various attacks detected by the respondents’ pool, is the use of toolkits (rootkits, meta-exploits, etc.) for the identification and exploitation of vulnerabilities
on the target system of the attack, with a 38.8\%, followed by the well-known malicious and unauthorized collection of information (social engineering, phishing, etc.) with 34.6\% 
and the use of malicious code and scripts with 34.3\%. The other considered technologies decrease with a few percentage points of difference between them. The impacts of the most 
critical attacks are analyzed for each attack type, as well as their possible motivations and the recovery times required by the most critical ones.
In the 2020 OAD survey, all these attacks’ characteristics vary for each type of attack, and the emerged results are described in the specific paragraphs dedicated to each attack 
type. At overall it emerges that:
\begin{itemize}
\item the impacts declared by the respondents are balanced between those irrelevant and/or easily resolvable with recovery in short time and at limited costs, and those very 
critical, which require expensive and long recovery actions, and that, in some cases, can cause business and customer loses; these two very different impact cases depend mainly 
on the security measures in place;

@@

\begin{figure}
	\centering
		\includegraphics[width=0.90\textwidth]{pictures/fig6.png}
		\caption{Average diffusion (percentage) of the attack techniques detected in the most serious attacks between 2019 - 1st quoter 2020 (multiple answers) \cite{oad20}}
		\label{fig:6}
\end{figure}
